### 2015-05-22 Harris_2015/manuscript-materials/draft.Rmd
done

Ecologists' intense interest in drawing inferences about species interactions
---especially competition---from presence-absence data has a long history
[@macarthur_population_1958; @diamond_island_1975; @connor_checkered_2013]. If
nontrophic species interactions are important drivers of community assembly patterns,
then we should expect to see their influence in our data sets. Despite decades
of work and several major controversies, however [@lewin_santa_1983;
@strong_ecological_1984; @gotelli_swap_2003], existing methods for detecting
competition's effects on community structure are unreliable
[@gotelli_empirical_2009]. More generally, it can be difficult to reason about
the complex web of direct and indirect interactions in real assemblages,
especially when these interactions occur
against a background of other ecological processes such as dispersal and
environmental filtering [@connor_checkered_2013]. For this reason, it isn't
always clear what kinds of patterns would even constitute evidence of
competition, as opposed to some other biological process or random sampling
error [@roughgarden_competition_1983; @lewin_santa_1983].

Most existing methods in this field compare the frequency with which two
putative competitors are observed to co-occur against the frequency that would
be expected if *all* species on the landscape were independent
[@strong_ecological_1984; @gotelli_empirical_2009]. Examining a species pair
against such a "null" background, however, rules out the possibility that the
overall association between two species could be driven by an outside force.
For example, even though the two shrub species in Figure 1 compete 
with one another for resources at a mechanistic level, they end up clustering
together on the landscape because they both grow best in areas that are not overshadowed
by trees. If this sort of effect is common, then significant deviations 
from independence will not---by themselves---provide convincing evidence of
species' direct effects on one another.

@@

In this paper, I introduce Markov networks [undirected graphical models
also known as Markov random fields; @murphy_machine_2012] as a framework for
understanding the landscape-level consequences of pairwise species
interactions, and for detecting them with observational data. Markov networks,
which generalize partial correlations to non-Gaussian data, have been used in
many scientific fields to model associations between various kinds of
"particles". For example, a well-studied network called the Ising model has played
an important role in our understanding of physics [where nearby particles tend
to align magnetically with one another; @cipra_introduction_1987]. In spatial
contexts, these models have been used to describe interactions between
adjacent grid cells [@harris_contact_1974; @gelfand_modelling_2005]. In
neurobiology, they have helped researchers determine which neurons are
connected to one another by modeling the structure in their firing patterns
[@schneidman_weak_2006]. Following recent work by @azaele_inferring_2010 and
@fort_statistical_2013, I suggest that ecologists could similarly treat
species as the interacting particles in this modeling framework. Doing so
would allow ecologists to simulate and study the landscape-level consequences
of arbitrary species interaction matrices, even when our observations are not
Gaussian. While ecologists explored some related approaches in the 1980's
[@whittam_species_1981], computational limitations had previously imposed
severe approximations that produced unintelligible
results [e.g. "probabilities" greater than one; @gilpin_factors_1982]. 
Now that it is computationally feasible to fit these models
exactly, the approach has become worth a second look.

The rest of the paper proceeds as follows. First, I discuss how Markov networks
work and how they can be used to simulate landscape-level data and to predict
the direct and indirect consequences of possible interaction matrices. Then,
using simulated data sets where the "true" ecological structure is known, I
compare this approach with several existing methods for detecting species
interactions. Finally, I discuss opportunities for extending the approach
presented here to larger problems in community ecology.

@@

Here, $\alpha_{i}$ is the amount that the presence of species $i$ contributes
to the log-probability of $\vec{y}$; it directly controls the prevalence of
species $i$. Similarly, $\beta_{ij}$ is the amount that the co-occurrence of
species $i$ and species $j$ contributes to the log-probability, and controls
how often the two species will be found together (Figure 2A, Figure 2B).
$\beta$ thus acts as an analog of the partial covariance, but for non-Gaussian
networks. Because the relative probability of a presence-absence vector
increases when positively-associated species co-occur and decreases when
negatively-associated species co-occur, the model tends to produce assemblages
that have many pairs of positively-associated species and relatively few pairs
of negatively-associated species (exactly as an ecologist might expect).

@@

Estimating the marginal relationships predicted by a Markov network is more
difficult than estimating conditional relationships, because doing so requires
absolute probability estimates. Turning the relative probability given by
Equation 1 into an absolute probability entails scaling by a *partition
function*, $Z(\alpha, \beta)$, which ensures that the probabilities of all
possible assemblages that could be produced by the model sum to one (bottom of
Figure 2B). Calculating $Z(\alpha, \beta)$ exactly, as is done in this paper,
quickly becomes infeasible as the number of species increases: with
$2^N$ possible assemblages of $N$ species the number of bookkeeping operations
required for exact inference quickly spirals exponentially into the billions.
Numerous techniques are available for working with Markov networks that keep
the computations tractable, either through analytic approximations
[@lee_learning_2012] or Monte Carlo sampling [@salakhutdinov_learning_2008],
but they are beyond the scope of this paper.

@@

The first set of simulated landscapes included the three competing
species shown in Figure 1. For each of 1000 replicates, I generated a
landscape with 100 sites by sampling exactly from a probability distribution
defined by the interaction coefficients in that figure (Appendix A). Each of the methods
described below (a Markov network, two correlation-based methods and a null modeling 
approach) was then evaluated on its ability to correctly infer that 
the two shrub species competed with one another, despite their frequent co-occurrence.

I also simulated a second set of landscapes with five, ten, or twenty
potentially-interacting species on landscapes composed of 20, 100, 500, or
2500 observed communities (24 replicate simulations for each combination; Appendix B).
These simulated data sets span the range from small, single-observer data
sets to large collaborative efforts such as the North American Breeding Bird
Survey. As described in Appendix B, I randomly drew the "true" coefficient
values for each replicate so that most species pairs interacted negligibly, a
few pairs interacted very strongly, and competition was three times more
common than facilitation. I then used Gibbs sampling to randomly generate
landscapes with varying numbers of species and sites via Markov chain Monte
Carlo (Appendix B). For half of the simulated landscapes, I treated each
species' $\alpha$ coefficient as a constant, as described above. For the other
half, I treated the $\alpha$ coefficients as linear functions of two abiotic
environmental factors that varied from location to location across the
landscape (Appendix B). The latter set of simulated landscapes provide an
important test of the methods' ability to distinguish co-occurrence patterns
that were generated from pairwise interactions among the observed species from
those that were generated by external forces like abiotic environmental
filtering. This task was made especially difficult because---as with most
analyses of presence-absence data for co-occurrence patterns---the inference
procedure did not have access to any information about the environmental or
spatial variables that helped shape the landscape [cf @connor_checkered_2013;
@blois_framework_2014].

@@

The first two interaction measures were the simple and partial covariances
between each pair of species' data vectors on the landscape
[@albrecht_spatial_2001]. Because partial covariances are undefined for
landscapes with perfectly-correlated species pairs, I used a regularized
estimate based on ridge regression [@van_wieringen_ridge_2014; i.e. linear
regression with a Gaussian prior]. For these analyses, I set the ridge
parameter to 0.2 divided by the number of sites on the landscape.

The third method, described in @gotelli_empirical_2009, involved simulating
possible landscapes from a null model that retains the row and column sums of
the original matrix [@strong_ecological_1984]. Using the default options in the
Pairs software described in @gotelli_empirical_2009, I simulated  the null
distribution of scaled C-scores (a test statistic describing the number of
*non*-co-occurrences between two species).  The software then calculated a
$Z$-statistic for each species pair using this null distribution. After
multiplying this statistic by $-1$ so that positive values corresponded to
facilitation and negative values corresponded to competition, I used it as
another estimate of species interactions.

***Method evaluation.***
For the first simulated landscape (three species), I kept the evaluation
simple and qualitative: any method that reliably determined that the two shrub
species were negatively associated passed; other methods failed.

For the larger landscapes, I rescaled the four methods' estimates using linear
regression through the origin so that they all had a consistent
interpretation. In each case, I regressed the "true" $\beta$ coefficient for
each species pair against the model's estimate, re-weighting the pairs so that
each landscape contributed equally to the rescaled estimate. For each 
estimate of a species pair's
interactions, I used this regression to calculate the squared error associated
with method that produced it.  Finally, I averaged these squared errors for
each combination of species richness, landscape size, statistical method, and 
presence/absence of environmental filtering across all 12 replicates; the mean
squared errors associated with these subsets of the data determined the proportion
of variance explained by each method under different conditions.

@@

***Three species.***
As shown in Figure 1, the marginal relationship between the two shrub species
was positive---despite their competition for space at a mechanistic level---
due to indirect effects of the dominant tree species. As a result, the
covariance method falsely reported positive associations 94% of the time
and the randomization-based null model falsely reported such associations 100%
of the time. The two methods for evaluating conditional relationships (Markov
networks and partial covariances), however, successfully controlled for the
indirect pathway via the tree species and each correctly identified the direct
negative interaction between the shrubs 94% of the time.

***Larger landscapes.***
The accuracy of the four evaluated methods varied substantially, depending
on the parameters that produced the simulated communities (Figure 3).
In general, however, there was a consistent ordering: the Markov network
explained 54% of the variance in the "true" parameters, followed by 
partial covariances (33%), and sample covariances (22%).

The null model scores initially explained 12% of the variation. After manually
reducing the value of one especially strong outlier ($Z=1004$, implying
$p<10^{-1000000}$), this increased to 17%.  Figure 3 reflects the adjusted
version of the results.

@@

The results presented above are very promising, as they show that Markov
networks can recover species' pairwise interactions from observational data,
even when direct interactions are largely overwhelmed by indirect effects
(Figure 1) or environmental effects (lower panels of Figure 3). For cases
where fitting a Markov network is infeasible, these results also indicate
that partial covariances---which can be computed straightforwardly by linear
regression---can often provide a surprisingly accurate approximation.
On the other hand, inferences may be much less reliable in cases where the "true"
interaction matrices include guild structure or trophic levels, which could
make the $\beta$ coefficients much more difficult to infer (particularly for
linear methods; @loh_structure_2013). In these cases, it may be necessary to
include prior information about the nature of species' interactions from 
natural history, experiments, or trait data. The rosalia package 
[@harris_rosalia_2015] has built-in mechanisms for incorporating 
this kind of information, if it can be expressed as a prior probability 
distribution on $\beta$ or a penalty on the likelihood.

Ecologists will also need experiments and natural history knowledge to pin
down the exact nature of the interactions identified by the network (e.g. whether
or not negative interactions are consumptive, or which species in a positively-associated
pair is facilitating the other), particularly when real pairs of species can reciprocally 
influence one another's occurrence probabilities in multiple ways simultaneously; the
$\beta$ coefficients in Markov networks have to reduce this complexity to a single
number. In short, partial correlations and Markov networks both prevent us 
from mistaking marginal associations for conditional ones, but they can't tell us the 
underlying biological mechanisms at work.

Despite these limitations, Markov networks have enormous potential to improve
ecological inferences.  For example, Markov networks provide a simple answer
to the question of how competition should affect a species' overall
prevalence, which was a major flash point for the null model debates in the
1980’s [@strong_ecological_1984]. Equation 1 can be used to calculate the
expected prevalence of a species in the absence of biotic influences
[$\frac{1}{1 + e^{-\alpha}}$; @lee_learning_2012]. Competition's effect on
prevalence in a Markov network can thus be calculated by subtracting this value
and the observed prevalence (cf Figure 2D).

Markov networks---particularly the Ising model for binary networks---have been studied 
in statistical physics for nearly a century [@cipra_introduction_1987], 
and the models' properties, 
capabilities, and limits are well-understood in a huge range of applications, 
from spatial statistics [@gelfand_modelling_2005] to neuroscience 
[@schneidman_weak_2006] to models of human behavior [@lee_statistical_2013].
Modeling species interactions using the same framework would thus allow
ecologists to tap into an enormous set of existing discoveries and techniques
for dealing with indirect effects, stability, and alternative stable states
(phase transitions).

This modeling approach is also highly
extensible, even when it is inconvenient to compute the likelihood exactly.
For example, the mistnet software package for joint species distribution modeling
[@harris_generating_2015] can fit *approximate* Markov networks to large
species assemblages (>300 species) while simultaneously modeling each species'
nonlinear response to the abiotic environment. Combining multiple ecological 
processes into a common model could help ecologists to disentangle different 
factors that can confound simpler co-occurrence analyses [cf @connor_checkered_2013]. 
Numerous other extensions are possible: similar networks
can be fit with a mix of discrete and continuous variables, for example
[@lee_learning_2012]. There are even methods [@whittam_species_1981;
@tjelmeland_markov_1998] that would allow the coefficient linking two species
in an interaction matrix to vary as a function of the abiotic environment or
of third-party species that tip the balance between facilitation and
exploitation [@bruno_inclusion_2003].

>>>>

If nontrophic species interactions, such as competition, are important drivers
of community assembly, then ecologists might expect to see their influence in
our data sets [@macarthur_population_1958; @diamond_island_1975]. Despite
decades of work and several major controversies, however [@lewin_santa_1983;
@strong_ecological_1984; @gotelli_swap_2003; @connor_checkered_2013], existing
methods for detecting competition's effects on community structure are
unreliable [@gotelli_empirical_2009], and thus important ecological processes
remain poorly understood. More generally, it can be difficult to reason about
the complex web of direct and indirect species interactions in real
assemblages, especially when these interactions occur against a background of
other ecological processes such as dispersal and environmental filtering
[@connor_checkered_2013]. For this reason, it isn't always clear what kinds of
patterns would even constitute evidence of competition, as opposed to some
other biological process or random sampling error
[@roughgarden_competition_1983; @lewin_santa_1983].

Most existing methods in this field compare the frequency with which two
putative competitors are observed to co-occur, versus the frequency that would
be expected if *all* species on the landscape were independent
[@strong_ecological_1984; @gotelli_empirical_2009]. Examining a species pair
against such a "null" background, however, rules out the possibility that the
overall association between two species could be driven by an outside force.
For example, even though the two shrub species in Figure 1 compete with one
another for resources at a mechanistic level, they end up clustering together
on the landscape because they both grow best in areas that are not
overshadowed by trees. If this sort of effect is common, then significant
deviations from independence will not generally provide convincing
evidence of species' direct effects on one another.

@@

In this paper, I introduce Markov networks [undirected graphical models also
known as Markov random fields; @murphy_machine_2012] as a framework for
understanding the landscape-level consequences of pairwise species
interactions, and for detecting them with observational data. Markov networks,
which generalize partial correlations to non-Gaussian data
[@lee_learning_2012; @loh_structure_2013], have been used in many scientific
fields for decades to model associations between various kinds of "particles".
For example, a well-studied network named the Ising model has played an
important role in our understanding of physics [where nearby particles tend to
align magnetically with one another; @cipra_introduction_1987]. In spatial
contexts, these models have been used to describe interactions between
adjacent grid cells [@harris_contact_1974; @gelfand_modelling_2005]. In
neurobiology, they have helped researchers determine which neurons are
connected to one another by modeling the structure in their firing patterns
[@schneidman_weak_2006]. Following recent work by @azaele_inferring_2010 and
@fort_statistical_2013, I suggest that ecologists could similarly treat
species as the interacting particles in the same modeling framework. Doing so
would allow ecologists to simulate and study the landscape-level consequences
of arbitrary species interaction matrices, even when our observations are not
Gaussian. While ecologists explored some related approaches in the 1980's
[@whittam_species_1981], computational limitations had previously imposed
severe approximations that produced unintelligible results [e.g.
"probabilities" greater than one; @gilpin_factors_1982].  Now that it is
computationally feasible to fit these models exactly, the approach has become
worth a second look.

The rest of the paper proceeds as follows. First, I discuss how Markov networks
work and how they can be used to simulate landscape-level data or to predict
the direct and indirect consequences of possible interaction matrices. Then,
using simulated data sets where the "true" ecological structure is known, I
compare this approach with several existing methods for detecting species
interactions. Finally, I discuss opportunities for extending the approach
presented here to larger problems in community ecology.

@@

Here, $\alpha_{i}$ is an intercept term determining the amount that 
the presence of species $i$ contributes
to the log-probability of $\vec{y}$; it directly controls the prevalence of
species $i$. Similarly, $\beta_{ij}$ is the amount that the co-occurrence of
species $i$ and species $j$ contributes to the log-probability; it controls
the probability that the two species will be found together (Figure 2A, Figure 2B).
$\beta$ thus acts as an analog of the partial covariance, but for non-Gaussian
networks. Because the relative probability of a presence-absence vector
increases when positively-associated species co-occur and decreases when
negatively-associated species co-occur, the model tends to produce assemblages
that have many pairs of positively-associated species and relatively few pairs
of negatively-associated species (exactly as an ecologist might expect).

@@

Estimating the marginal relationships predicted by a Markov network is more
difficult than estimating conditional relationships, because doing so requires
absolute probability estimates. Turning the relative probability given by
Equation 1 into an absolute probability entails scaling by a *partition
function*, $Z(\alpha, \beta)$, which ensures that the probabilities of all
possible assemblages that could be produced by the model sum to one (bottom of
Figure 2B). Calculating $Z(\alpha, \beta)$ exactly, as is done in this paper,
quickly becomes infeasible as the number of species increases: with
$2^N$ possible assemblages of $N$ species, the number of bookkeeping operations
required for exact inference spirals exponentially into the billions and beyond.
Numerous techniques are available for working with Markov networks that keep
the computations tractable, e.g. via analytic approximations
[@lee_learning_2012] or Monte Carlo sampling [@salakhutdinov_learning_2008],
but they are beyond the scope of this paper.

@@

The first set of simulated landscapes included the three competing species
shown in Figure 1. For each of 1000 replicates, I generated a landscape with
100 sites by sampling exactly from a probability distribution defined by the
interaction coefficients in that figure (Appendix A). Each of the methods
described below (a Markov network, two correlation-based methods and a null
model) was then evaluated on its ability to correctly infer that the two
shrub species competed with one another, despite their frequent co-occurrence.

I also simulated a second set of landscapes with five, ten, or twenty
potentially-interacting species on landscapes composed of 20, 100, 500, or
2500 observed communities (24 replicate simulations for each combination; Appendix B).
These simulated data sets span the range from small, single-observer data
sets to large collaborative efforts such as the North American Breeding Bird
Survey. As described in Appendix B, I randomly drew the "true" coefficient
values for each replicate so that most species pairs interacted negligibly, a
few pairs interacted very strongly, and competition was three times more
common than facilitation. I then used Gibbs sampling to randomly generate
replicate landscapes with varying numbers of species and sites (Appendix B). 
For half of the simulated landscapes, I treated each
species' $\alpha$ coefficient as a constant, as described above. For the other
half, I treated the $\alpha$ coefficients as linear functions of two abiotic
environmental factors that varied from location to location across the
landscape (Appendix B). The latter set of simulated landscapes provide an
important test of the methods' ability to distinguish co-occurrence patterns
that were generated from pairwise biotic interactions from
those that were generated by external forces like abiotic environmental
filtering. This task was made especially difficult because---as with most
analyses of presence-absence data for co-occurrence patterns---the inference
procedure did not have access to any information about the environmental or
spatial variables that helped shape the landscape [cf @connor_checkered_2013;
@blois_framework_2014].

@@

The first two interaction measures were the sample covariances and the 
partial covariances between each pair of species' data vectors on the landscape
[@albrecht_spatial_2001]. Because partial covariances are undefined for
landscapes with perfectly-correlated species pairs, I used a regularized
estimate based on ridge regression [@van_wieringen_ridge_2014; i.e. linear
regression with a Gaussian prior]. For these analyses, I set the ridge
parameter to 0.2 divided by the number of sites on the landscape.

The third method, described in @gotelli_empirical_2009, involved simulating
possible landscapes from a null model that retains the row and column sums of
the original matrix [@strong_ecological_1984]. Using the default options in the
Pairs software described in @gotelli_empirical_2009, I simulated  the null
distribution of scaled C-scores (a test statistic describing the number of
*non*-co-occurrences between two species). The software then calculated a
$Z$ statistic for each species pair using this null distribution. After
multiplying this statistic by $-1$ so that positive values corresponded to
facilitation and negative values corresponded to competition, I used it as
another estimate of species interactions.

***Method evaluation.***
I evaluated each method qualitatively on the simulated landscapes based on
Figure 1: any method that reliably determined that the two shrub
species were negatively associated passed; other methods failed.

For the larger landscapes, I rescaled the four methods' estimates using linear
regression through the origin so that they all had a consistent
interpretation. In each case, I regressed the "true" $\beta$ coefficient for
each species pair against the model's estimate, re-weighting the pairs so that
each landscape contributed equally to the rescaled estimate. For each  method,
I calculated the squared deviations between each species pair's rescaled
estimate and the "true" interactions used to generate the assemblages.
Finally, I  averaged these squared errors for each combination of species
richness, landscape size, statistical method, and presence/absence of
environmental filtering across all 12 replicates; the mean squared errors
associated with these subsets of the data determined the proportion of
variance explained by each method under different conditions.

@@

***Three species.***
As shown in Figure 1, the marginal relationship between the two shrub species
was positive---despite their competition for space at a mechanistic level---
due to indirect effects of the dominant tree species. As a result, the
covariance method falsely reported positive associations on 94% of simulated landscapes,
and the randomization-based null model falsely reported such associations 100%
of the time. The two methods for evaluating conditional relationships (Markov
networks and partial covariances), however, successfully controlled for the
indirect pathway via the tree species and each correctly identified the direct
negative interaction between the shrubs 94% of the time.

***Larger landscapes.***
The accuracy of the four evaluated methods varied substantially, depending
on the parameters that produced the simulated communities (Figure 3).
In general, however, there was a consistent ordering: overall, the Markov network
explained 54% of the variance in the "true" parameters, followed by 
partial covariances (33%), and sample covariances (22%).

The null model scores initially explained only 12% of the variation. After
manually reducing the value of one especially strong outlier ($Z=1004$,
implying $p<10^{-1000000}$), this increased to 17% (Appendix B).  Figure 3
reflects the adjusted version of the results.

@@

The results presented above show that Markov networks, unlike a common null
modeling approach, can recover species' pairwise interactions from
observational data. The Markov networks were successful even when direct
interactions are largely overwhelmed by indirect effects (Figure 1) or
environmental effects (lower panels of Figure 3). For cases where fitting a
Markov network is computationally infeasible, these results also indicate that
partial covariances---which can be computed straightforwardly by linear
regression---can often provide a surprisingly useful approximation. The
success of the partial correlation's success on simulated data may not 
carry over to real data sets, however; @loh_structure_2013 suggest that the linear 
approximations may be less reliable in cases where the true interaction matrix 
contains more structure (e.g. guilds or trophic levels). On the other hand,
if ecologists are familiar enough with the natural history of their study
systems to describe this kind of structure as a prior distribution on the
parameters or as a penalty on the likelihood, then real-world results might
be even better than those shown in Figure 3.

Ecologists will also need natural history to pin down the exact nature of the
interactions identified by a network model (e.g. which species in a
positively-associated pair is facilitating the other),  particularly when real
pairs of species can reciprocally influence one another in multiple ways
simultaneously [@bruno_inclusion_2003]; the $\beta$ coefficients in Markov
networks have to reduce this complexity  to a single number. In short, partial
correlations and Markov  networks both help prevent us from mistaking marginal
associations for  conditional ones, but they can't tell us the underlying
biological mechanisms at work.

Despite these limitations, Markov networks have enormous potential to improve
ecological inferences. For example, Markov networks provide a simple answer
to the question of how competition should affect a species' overall
prevalence, which was a major flash point for the null model debates in the
1980’s [@roughgarden_competition_1983; @strong_ecological_1984]. Equation 
1 can be used to calculate the expected prevalence of a species in the 
absence of biotic influences [$\frac{1}{1 + e^{-\alpha}}$; @lee_learning_2012]. 
Competition's effect on prevalence in a Markov network can then be calculated 
by subtracting this value and the observed prevalence (cf Figure 2D).

Markov networks---particularly the Ising model for binary networks---have been studied 
in statistical physics for nearly a century [@cipra_introduction_1987], 
and the models' properties, 
capabilities, and limits are well-understood in a huge range of applications, 
from spatial statistics [@gelfand_modelling_2005] to neuroscience 
[@schneidman_weak_2006] to models of human behavior [@lee_statistical_2013].
Modeling species interactions using the same framework would thus allow
ecologists to tap into an enormous set of existing discoveries and techniques
for dealing with indirect effects, stability, and alternative stable states
(i.e. phase transitions; @cipra_introduction_1987).

This modeling approach is also highly
extensible, even when it is inconvenient to compute the likelihood exactly.
For example, the mistnet software package for joint species distribution modeling
[@harris_generating_2015] can fit *approximate* Markov networks to large
species assemblages (>300 species) while simultaneously modeling each species'
nonlinear response to the abiotic environment. Combining multiple ecological 
processes into a common model could help ecologists to disentangle different 
factors that can confound simpler co-occurrence analyses [cf @connor_checkered_2013]. 
Numerous other extensions are possible: Markov networks
can be fit with a mix of discrete and continuous variables, for example
[@lee_learning_2012]. There are even methods [@whittam_species_1981;
@tjelmeland_markov_1998] that would allow the coefficient linking two species
in an interaction matrix to vary as a function of the abiotic environment or
of third-party species that tip the balance between facilitation and
exploitation [@bruno_inclusion_2003].
